---
title: "Testing predicted temperature for exercice 3 - Week 4"
author: "Caio Geraldes"
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
---

```{r echo=FALSE, include=FALSE}
library(tidyverse)
library(rethinking)
library(dagitty)
```

# Loading data

```{r}
data("cherry_blossoms")
d.cherry <- (cherry_blossoms)
d.cherry <- d.cherry[complete.cases(d.cherry$temp,d.cherry$doy),]
```

# Modeling

## DAG, causal model

```{r echo=FALSE}
cherry.dag <- dagitty('dag {
D [outcome,pos="-0.750,-0.500"]
T [exposure,pos="-1.000,-0.500"]
T_l [adjusted,pos="-1.000,-0.250"]
T_u [adjusted,pos="-1.000,-0.750"]
Y [pos="-1.250,-0.500"]
T -> D
T <-> T_l
T <-> T_u
T_l -> D
T_u -> D
Y -> T
Y -> T_l
Y -> T_u
}')
drawdag(cherry.dag)
```

## Temperature Model A: polynomial regression (on date)

Model: $temp \sim \alpha + \sum_{k=1}^{5}{w_k*year^k}$

Prior:

$temp_i \sim \text{Normal}(\mu_i, \sigma)\\\mu_i = \alpha + \sum_{k=1}^{5}{w_k*year^k}\\\alpha \sim \text{Normal}(100,10)\\w_k \sim \text{Normal}(0, 10)\\\sigma \sim \text{Exponential}(1)$

```{r}
m.temp <- quap(
    alist(
        temp ~ dnorm( mu , sigma ) ,
        mu <- a + w1*year + w2*year^2 + w3*year^3 + w4*year^4 + w5*year^5,
        a ~ dnorm(1,1),
        w1 ~ dnorm(0,10),
        w2 ~ dnorm(0,10),
        w3 ~ dnorm(0,10),
        w4 ~ dnorm(0,10),
        w5 ~ dnorm(0,10),
        sigma ~ dexp(1)
    ), data=list( temp=d.cherry$temp , year=standardize(d.cherry$year) ) 
    )
```

```{r}
precis(m.temp, depth=2)
precis_plot(precis(m.temp, depth=2))
```

```{r}
mu <- link(m.temp)
mu.mean <- apply(mu,2,mean)
mu.min <- apply(mu,2,PI,prob=0.99)["1%",]
mu.max <- apply(mu,2,PI,prob=0.99)["100%",]

w_sim <- sim(m.temp)
w_sim.min <- apply(w_sim,2,PI)['5%',]
w_sim.max <- apply(w_sim,2,PI)['94%',]

d.cherry %>% ggplot(aes(x=year, y=temp)) +
    geom_ribbon(aes(ymax=w_sim.max, ymin=w_sim.min), fill='grey', alpha=0.4) +
    geom_ribbon(aes(ymax=mu.max, ymin=mu.min), fill='blue', alpha=.3) +
    geom_point(alpha=.6)
```

## DOY Model SLR A: simple linear regression, unstratified

Model: $doy \sim \alpha + \beta_t * temperature$

Prior:

$doy_i \sim \text{Normal}(\mu_i, \sigma)\\\mu_i = \alpha + \beta_t * temperature_i\\\alpha \sim \text{Normal}(100,10)\\\beta_t \sim \text{Normal}(0, 10)\\\sigma \sim \text{Exponential}(1)$

```{r echo=TRUE, results='hide'}
m.slr.a <- ulam(
    alist(
        doy ~ dnorm(mu, sigma),
        mu <- alpha + beta_t * temp,
        alpha ~ dnorm(100,10),
        beta_t ~ dnorm(0,10),
        sigma ~ dexp(1)
        ),
    data=d.cherry, chains=4, cores=4, log_lik=TRUE)
```

```{r}
precis(m.slr.a)
precis_plot(precis(m.slr.a))
```


# Predictions

```{r}
z <- (2050 - mean(d.cherry$year)) / sd(d.cherry$year)
temp_sim <- sim(m.temp, data = list(year=z))

temp_sim.dens <- with(density(temp_sim), tibble(x, y))

doy_sim_2050 <- flatten_dbl(as.list(sim(m.slr.a, data = list(temp=temp_sim))))
doy.slr.2050.dens <- with(density(doy_sim_2050), tibble(x, y))
doy.slr.2050.dens$model <- "Pred. temp. for 2050"

doy_sim_9 <- flatten_dbl(as.list(sim(m.slr.a, data = list(temp=9))))
doy.slr.9.dens <- with(density(doy_sim_9), tibble(x, y))
doy.slr.9.dens$model <- "Temp = 9 degrees"
```



```{r}
(hpdi.2050 <- HPDI(doy_sim_2050))
(hpdi.9 <- HPDI(doy_sim_9))

doy.data.dens <- with(density(d.cherry$doy), tibble(x, y))
doy.data.dens$model <- "Data"
(hpdi.data <- HPDI(d.cherry$doy))

doy.slr.2050.dens %>% ggplot(aes(x,y)) +
    geom_line() +
    geom_area(data=subset(doy.slr.2050.dens, x > hpdi.2050[1] & x < hpdi.2050[2]), aes(fill=model), alpha=0.5) +
    geom_line(aes(x,y), data=doy.data.dens) +
    geom_area(data=subset(doy.data.dens, x > hpdi.data[1] & x < hpdi.data[2]), aes(fill = model), alpha=0.5) +
    geom_line(aes(x,y), data=doy.slr.9.dens) +
    geom_area(data=subset(doy.slr.9.dens, x > hpdi.9[1] & x < hpdi.9[2]), aes(fill = model), alpha=0.5) +
    xlab('Date of blossom') +
    ylab('density') +
    labs(title='Contrast between current date of blossom with simulations') +
    scale_fill_brewer("",palette = 'Dark2')
```